Experiences with Planning for Natural Language Generation

### 2009-05-28 trunk/Papers/ci-spark/nlg-as-planning.tex
gen as planning

The task of generating natural language from semantic representations
is typically split up into two parts: the \emph{domain planning} task,
which selects the information that is to be conveyed and structures it
into sentence-sized chunks, and the \emph{sentence generation} tasks,
which then translates each of these chunks into natural language
sentences. The sentence generation task is often split up into two
parts of its own---the \emph{sentence planning} task, which enriches
the input by e.g.\ determining how to refer to objects and selects
some lexical material, and the \emph{surface realization} task, which
maps the enriched meaning representation into a sentence using a
grammar. The chain of domain planning, sentence planning, and surface
realization is sometimes called the ``NLG pipeline''
\cite{reiter00building}.

Viewing generation as a planning problem has a long tradition in the
NLG literature, especially in discourse planning. \todo{Sketch
  Perrault et al here. Say something very brief about Appelt, Hovy,
  Moore. Say why all this didn't really work.}

Most recently, interest in generation as planning has been revived for
NLG tasks on various levels of the NLG pipeline. Below, we will look
in some detail at \cite{KolSto07}'s recasting of sentence generation
as planning; furthermore, \cite{Steedman-Petrick:07} and
\cite{benotti08b} have applied planning to \todo{something}. Notice
that although all of these approaches encode ``communicative actions''
as planning operators, they still solve fundamentally different
problems. For instance, Perrault et al.\ tackled discourse planning
with planning operators that encode speech acts, each of which
represents the meaning of a whole sentence and models the contribution
of this sentence to the intentional structure of the entire
discourse. By contrast, Koller and Stone solve a sentence generation
problem using planning operators that encode utterances of single
words and model the contribution of each word to the grammatical
structure of the sentence. Whether these different levels of ``NLG as
planning'' could be combined into a single planning problem, and
whether this would make computational sense, is an open question.

Problems in NLG at all levels of description can become
computationally complex very easily. For instance, \citet{KolStr02}
showed that even surface realization is NP-complete under reasonable
assumptions. This makes it necessary to investigate methods for
guiding the search, which makes modern planning techniques a promising
approach to NLG. The point of this paper is to investigate whether
existing planners fulfill this promise of efficiency for the
application to NLG.

>>>>

The task of generating natural language from semantic representations
(NLG) is typically split up into two parts: the \emph{discourse
  planning} task, which selects the information that is to be conveyed
and structures it into sentence-sized chunks, and the \emph{sentence
  generation} tasks, which then translates each of these chunks into
natural language sentences. The sentence generation task is often
split up into two parts of its own---the \emph{sentence planning}
task, which enriches the input by e.g.\ determining how to refer to
objects and selects some lexical material, and the \emph{surface
  realization} task, which maps the enriched meaning representation
into a sentence using a grammar. The chain of domain planning,
sentence planning, and surface realization is sometimes called the
``NLG pipeline'' \cite{reiter00building}.

Viewing generation as a planning problem has a long tradition in the
NLG literature.  \citet{perrault80} presented an approach to discourse
planning in which the planning operators represent individual speech
acts such as ``request'' and ``inform''. This idea was later worked
out further, e.g.\ by \citet{young94dpocl}. On the other hand,
researchers such as \citet{appelt:planning} and \citet{hovy88} used
techniques from hierarchical planning to expand a high-level plan
consisting of speech acts into more detailed specifications of
individual sentences. Their systems also covered some aspects of
sentence planning. However, these systems used very expressive logics,
which were designed to reason about beliefs and intentions, to
represent the planning state and the planning operators. Furthermore,
most of them used ad-hoc planning algorithms with rather naive search
strategies, which did not scale to realistic inputs. As a consequence,
the NLG-as-planning approach was mostly marginalized throughout the
1990s.

However, there has recently been a string of publications by various
authors with renewed interest in the generation-as-planning approach,
motivated by the recent development of efficient and expressive
planners. Below (Section~\ref{sec:domain-crisp}) we will look in some
detail at \cite{KolSto07}'s approach to sentence generation (i.e., the
sentence planning and surface realization modules of the pipeline) as
planning. In addition, \citet{Steedman-Petrick:07} revisit the
analysis of indirect speech acts with modern planning technology;
\citet{benotti08b} uses planning to explain the accommodation of
presuppositions; and
\citet{brenner08:_contin_multiag_plann_approac_to_situat_dialog} use
multi-agent planning to model the joint problem-solving behavior of
agents in a situated dialogue. These approaches focus on different
issues than the 1980's NLG-as-planning literature. But the main
difference is that they apply \emph{existing}, well-understood
planning approaches to linguistic problems because these offer new
modeling tools, and because they hope that modern planners can solve
the hard search problems that arise in NLG \citep{KolStr02}
efficiently. The point of this paper is to investigate whether
existing planners fulfill this latter promise.
